---
title: "Getting Started with cliomicplot"
author: "Lucas VHH Tran"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    mathjax: null
vignette: >
  %\VignetteIndexEntry{Getting Started with cliomicplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center"
)
```

## Overview

**cliomicplot** is an R package for creating high-quality, publication-ready visualizations
for clinical and multi-omics data analysis. Built on **ggplot2**, it provides 16 specialized
plot types, 10 journal-specific themes, and 55+ color palettes — all through a clean,
formula-based API.

This vignette walks you through the core concepts and basic usage.

## Installation

```{r eval=FALSE}
# From source
remotes::install_local("path/to/cliomicplot")

# Dependencies
install.packages(c(
  "ggplot2", "scales", "survival", "survminer",
  "RColorBrewer", "reshape2", "ggrepel",
  "patchwork", "ggpubr", "ggtext", "circlize"
))
```

```{r}
library(cliomicplot)
```

---

## The Formula Interface

cliomicplot uses a formula-based API: `y ~ x | group`.

```{r}
# Basic scatter with grouping
cliplot(Sepal.Length ~ Petal.Length | Species, data = iris)
```

The `| group` part is optional. Every plot automatically gets:

- Proper axis labels derived from variable names
- A legend when grouping is present
- A color palette applied to groups

---

## Choosing a Plot Type

Pass `type = "..."` or a `type_*()` function:

```{r}
# Boxplot with built-in statistical test
cliplot(Sepal.Length ~ Species, data = iris,
        type = "boxplot",
        stat.test = "kruskal.test",
        stat.label = "p.signif",
        palette = "jco")
```

When `type` is omitted, cliomicplot guesses based on data types:

| x variable | y variable | Auto-detected type |
|---|---|---|
| numeric | numeric | `"points"` (scatter) |
| factor/character | numeric | `"boxplot"` |
| factor/character | _(none)_ | `"barplot"` |
| numeric | _(none)_ | `"histogram"` |

---

## Statistical Annotations

Built-in statistical test support — no extra packages needed:

```{r}
# t-test with p-value formatting
cliplot(len ~ dose, data = ToothGrowth,
        type = "boxplot",
        stat.test = "t.test",
        stat.label = "p.format",
        palette = "jco")
```

Supported tests: `"t.test"`, `"wilcox.test"`, `"anova"`, `"kruskal.test"`.

Supported label formats: `"p.format"` (e.g., `p = 0.023`), `"p.signif"` (e.g., `*`, `**`), `"p.adj"`.

---

## Theming

### Persistent themes

Set a theme that applies to **all** subsequent plots:

```{r}
clitheme("nature")
cliplot(mpg ~ wt, data = mtcars)
```

```{r}
# The theme persists
cliplot(len ~ dose, data = ToothGrowth, type = "boxplot")
clitheme()  # reset to default
```

### Ephemeral (per-plot) themes

Apply a theme to **one plot only**:

```{r}
cliplot(mpg ~ wt, data = mtcars, theme = "nature")
cliplot(Sepal.Length ~ Species, data = iris, type = "boxplot", theme = "dark")
```

### Available themes

| Theme | Style |
|---|---|
| `cli_bw` | Clean black-and-white (default) |
| `cli_classic` | Classic with axis lines |
| `cli_minimal` | Minimal and modern |
| `nature` | Nature Publishing Group |
| `science` | Science / AAAS |
| `nejm` | New England Journal of Medicine |
| `lancet` | The Lancet |
| `cell` | Cell Press |
| `broadsheet` | Publication print style |
| `dark` | Dark background (presentations) |

---

## Color Palettes

cliomicplot ships with 55+ palettes in a unified system:

```{r}
# List all palettes (first 20 shown)
head(cli_palette_list(), 20)
```

```{r}
# Use any palette by name
cliplot(Sepal.Length ~ Petal.Length | Species, data = iris,
        palette = "jama")

cliplot(Sepal.Length ~ Petal.Length | Species, data = iris,
        palette = "cosmic")
```

### Standalone ggplot2 scale usage

All palettes work as standalone scales for any ggplot:

```{r}
library(ggplot2)

ggplot(iris, aes(Sepal.Length, Petal.Length, color = Species)) +
  geom_point(size = 3) +
  palette_scale("npg", "color") +
  theme_bw()
```

---

## Markdown Text Rendering

Add **bold**, *italic*, colored, subscript/superscript text to labels:

```{r}
cliplot(mpg ~ wt, data = mtcars,
        title    = "**Mileage** vs *Weight* in mtcars",
        subtitle = "n = 32 vehicles, <span style='color:#E64B35'>1974 Motor Trend</span>",
        caption  = "*Motor Trend* (1974)") +
  cli_markdown()
```

### Selective markdown

```{r}
# Only the title renders markdown
cliplot(mpg ~ wt, data = mtcars,
        title = "**Bold Title**",
        xlab  = "Weight (1000 lbs)",    # plain text
        ylab  = "Miles / Gallon") +      # plain text
  cli_markdown("title")
```

---

## Global Parameters with `clipar()`

Set defaults that apply to all subsequent plots:

```{r}
# View current parameters (first 10)
names(clipar())[1:10]
```

```{r}
# Set defaults
clipar(palette.qualitative = "nejm")
clipar(stat.test = "wilcox.test")
clipar(file.width = 8, file.height = 6, file.res = 600)

# Query
clipar("palette.qualitative")
```

---

## Saving to File

Save plots directly from `cliplot()` in one call. Supports pdf, png, jpg, svg,
and tiff:

```{r eval=FALSE}
cliplot(Sepal.Length ~ Petal.Length | Species, data = iris,
        file   = "iris_scatter.pdf",
        width  = 6,
        height = 5)

# PNG with high DPI
cliplot(mpg ~ wt, data = mtcars,
        file = "mpg_wt.png", width = 8, height = 6)
```

Set global defaults for file output:

```{r eval=FALSE}
clipar(file.width = 8, file.height = 6, file.res = 600)
cliplot(Sepal.Length ~ Petal.Length | Species, data = iris,
        file = "fig1_highres.pdf")
```

---

## Next Steps

- **`vignette("oncology", package = "cliomicplot")`** — End-to-end oncology analysis
- **`vignette("multiomics", package = "cliomicplot")`** — Multi-omics exploratory analysis
- **`vignette("themes-palettes", package = "cliomicplot")`** — Deep dive into themes & palettes
