---
title: "Multi-Omics Exploratory Analysis with cliomicplot"
author: "Lucas VHH Tran"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    mathjax: null
vignette: >
  %\VignetteIndexEntry{Multi-Omics Exploratory Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7.5,
  fig.height = 5.5,
  fig.align  = "center"
)
```

## Overview

This vignette demonstrates a multi-omics exploratory analysis workflow using
**cliomicplot**. We'll generate four figures commonly found in omics publications:

1. **Figure 1** — PCA with confidence ellipses
2. **Figure 2** — Correlation heatmap with clustering and significance
3. **Figure 3** — Volcano plot for differential expression
4. **Figure 4** — MA plot for DE quality assessment

```{r}
library(cliomicplot)
clipar(stat.test = NULL)  # disable auto stat annotations
```

---

## Figure 1: PCA with Confidence Ellipses

Principal Component Analysis (PCA) is the workhorse of omics data exploration.
cliomicplot can compute PCA from raw data or accept pre-computed results:

### From raw data

```{r}
clitheme("nature")

cliplot(iris[, 1:4],
        type = type_pca(
          pc_x          = 1,
          pc_y          = 2,
          center        = TRUE,
          scale.        = TRUE,
          add_ellipse   = TRUE,
          ellipse_level = 0.95,
          point_size    = 2.5,
          label_samples = FALSE
        ),
        by       = iris$Species,
        palette  = "npg",
        stat.test = NULL,
        title    = "**Figure 1** — PCA of Iris Morphometric Data",
        subtitle = "PC1 (72.9%) vs. PC2 (22.9%), 95% CI ellipses") +
  cli_markdown()
```

The axis labels automatically include variance explained percentages, and
confidence ellipses are drawn around each group.

### From pre-computed PCA

When you already have PCA results (e.g., from `prcomp()`, `DESeq2`, or another
pipeline), cliomicplot accepts the scores directly:

```{r}
pca_res <- prcomp(iris[, 1:4], center = TRUE, scale. = TRUE)
scores  <- as.data.frame(pca_res$x)
scores$Species <- iris$Species

cliplot(PC2 ~ PC1, data = scores, by = scores$Species,
        type      = "points",
        stat.test = NULL,
        palette   = "d3_category10",
        title     = "**PCA Scores** — Pre-computed by prcomp()",
        theme     = "science") +
  cli_markdown()
```

### PCA with sample labels

```{r}
cliplot(iris[, 1:4],
        type = type_pca(
          add_ellipse   = FALSE,
          label_samples = TRUE,
          label_size    = 2.5,
          point_size    = 2
        ),
        by      = iris$Species,
        palette = "jco",
        stat.test = NULL,
        title   = "**PCA with Sample Labels**") +
  cli_markdown()
```

---

## Figure 2: Correlation Matrix

Correlation analysis with significance stars and hierarchical clustering:

```{r}
clitheme("cli_minimal")

cliplot(mtcars[, 1:7],
        type = type_correlation(
          method    = "pearson",
          type      = "lower",
          add_coef  = TRUE,
          coef_size = 3.5,
          cluster   = TRUE,
          sig_level = 0.001
        ),
        stat.test = NULL,
        title = "**Figure 2** — Pearson Correlation Matrix (mtcars)",
        subtitle = "Lower triangle | *** p < 0.001") +
  cli_markdown()
```

### Correlation options

```{r}
# Spearman correlation, upper triangle, no coefficients
cliplot(mtcars[, 1:5],
        type = type_correlation(
          method   = "spearman",
          type     = "upper",
          add_coef = FALSE,
          cluster  = FALSE
        ),
        title = "**Spearman Correlation** — Upper Triangle") +
  cli_markdown()
```

```{r}
# Full matrix with all significance levels
cliplot(mtcars[, 1:6],
        type = type_correlation(
          method    = "pearson",
          type      = "full",
          add_coef  = TRUE,
          cluster   = TRUE,
          sig_level = 0.05
        ),
        title = "**Full Correlation Matrix** with Significance Stars") +
  cli_markdown()
```

---

## Figure 3: Volcano Plot — Differential Expression

The volcano plot is the canonical visualization for differential expression analysis.
We'll generate realistic simulated data:

```{r}
set.seed(42)

# Generate 8000 "genes" — most are null
n_genes <- 8000
de_data <- data.frame(
  logFC   = rnorm(n_genes, 0, 0.7),
  PValue  = 10^-rnorm(n_genes, 2.5, 1.8),
  row.names = paste0("GENE", 1:n_genes)
)

# Inject 200 "true" DE genes
sig_idx <- sample(1:n_genes, 200)
de_data$logFC[sig_idx]  <- rnorm(200, ifelse(runif(200) > 0.5, 2.0, -2.0), 0.5)
de_data$PValue[sig_idx] <- 10^-runif(200, 5, 15)

cat(sprintf("Genes tested: %d\n", n_genes))
cat(sprintf("DE genes (|log2FC| > 1, p < 0.05): %d\n",
            sum(abs(de_data$logFC) > 1 & de_data$PValue < 0.05)))
```

```{r}
clitheme("cell")

cliplot(-log10(PValue) ~ logFC, data = de_data,
        type = type_volcano(
          pval_cutoff  = 0.01,
          fc_cutoff    = 1,
          label_genes  = "significant",
          max_overlaps = 25,
          point_alpha  = 0.5
        ),
        title    = "**Figure 3** — Volcano Plot of Differential Expression",
        subtitle = "FDR < 0.01, |log<sub>2</sub>FC| > 1",
        caption  = paste0("Top labeled genes; ", n_genes, " probes tested")) +
  cli_markdown()
```

### Volcano plot color scheme

| Category | Condition | Default Color |
|---|---|---|
| **Up** (red) | $p < p_\text{cutoff}$ and $\log_2(\text{FC}) > \text{fc\_cutoff}$ | `#E64B35` |
| **Down** (blue) | $p < p_\text{cutoff}$ and $\log_2(\text{FC}) < -\text{fc\_cutoff}$ | `#4DBBD5` |
| **NS** (grey) | All other genes | `grey70` |

### Customizing the volcano

```{r}
# Tighter thresholds, different colors
cliplot(-log10(PValue) ~ logFC, data = de_data,
        type = type_volcano(
          pval_cutoff  = 0.001,
          fc_cutoff    = 1.5,
          label_genes  = NULL,       # no labels
          up_color     = "#BC3C29",  # NEJM red
          down_color   = "#0072B5",  # NEJM blue
          ns_color     = "grey85",
          point_size   = 1.5,
          point_alpha  = 0.4
        ),
        palette = "volcano",
        title   = "**Volcano Plot** — Stringent Thresholds",
        subtitle = "p < 0.001, |log<sub>2</sub>FC| > 1.5") +
  cli_markdown()
```

---

## Figure 4: MA Plot — DE Quality Check

The MA (Mean-Average) plot complements the volcano by showing the relationship
between fold-change and mean expression level:

```{r}
set.seed(123)
ma_data <- data.frame(
  baseMean       = 10^runif(3000, 0, 5),
  log2FoldChange = rnorm(3000, 0, 0.6),
  padj           = runif(3000)
)
# Add "real" DE genes
ma_data$log2FoldChange[sample(1:3000, 150)] <- rnorm(150, 2, 0.5)
ma_data$log2FoldChange[sample(1:3000, 150)] <- rnorm(150, -2, 0.5)

clitheme("science")

cliplot(log2FoldChange ~ baseMean, data = ma_data,
        type = type_ma(
          pval_cutoff = 0.05,
          add_loess   = TRUE,
          loess_color = "#3C5488",
          sig_color   = "#E64B35",
          ns_color    = "grey70",
          point_size  = 1.2,
          point_alpha = 0.4
        ),
        title    = "**Figure 4** — MA Plot",
        subtitle = "DE genes highlighted (padj < 0.05), LOESS trend") +
  cli_markdown()
```

The MA plot shows:
- **y-axis**: $\log_2$ fold change (biological effect size)
- **x-axis**: Mean normalized counts (expression level, log₁₀ scale)
- **Red points**: Significant DE genes
- **Blue curve**: LOESS trend line (checks for intensity-dependent bias)
- **Horizontal dashed line**: $\log_2\text{FC} = 0$ (no change)

---

## Heatmap: Clustered Expression Matrix

For completeness, here's a clustered heatmap workflow:

```{r}
set.seed(789)
expr_mat <- matrix(rnorm(500), nrow = 50)
rownames(expr_mat) <- paste0("Gene_", 1:50)
colnames(expr_mat) <- paste0("Sample_", sprintf("%02d", 1:10))

# Column annotations
ann_col <- data.frame(
  Condition = rep(c("Control", "Treatment"), each = 5),
  Batch     = rep(c("A", "B"), 5),
  row.names = colnames(expr_mat)
)

# Annotation color mapping
ann_colors <- list(
  Condition = c(Control = "#4DBBD5", Treatment = "#E64B35"),
  Batch     = c(A = "#E18727", B = "#3C5488")
)
```

```{r, eval=requireNamespace("ComplexHeatmap", quietly=TRUE)}
cliplot(expr_mat,
        type = type_heatmap(
          scale             = "row",
          cluster_rows      = TRUE,
          cluster_cols      = TRUE,
          annotation_col    = ann_col,
          annotation_colors = ann_colors,
          color_low         = "#4575B4",
          color_mid         = "white",
          color_high        = "#D73027",
          fontsize          = 9
        ))
```

> **Note:** The heatmap uses `ComplexHeatmap` when available, falling back to
> `pheatmap`. Install with `install.packages("ComplexHeatmap")` for the best experience.

---

## Summary

In this vignette we produced four publication-ready omics figures:

1. **PCA** — Dimensionality reduction with confidence ellipses
2. **Correlation Matrix** — Pairwise correlations with hierarchical clustering
3. **Volcano Plot** — Differential expression significance vs. fold-change
4. **MA Plot** — Fold-change vs. mean expression with LOESS trend

Reset global settings:

```{r}
clitheme()
```
