---
title: "Multi-Omics Exploratory Analysis with cliomicplot"
author: "Lucas VHH Tran"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    mathjax: null
    css: reportdown.css
    self_contained: true
    includes:
      in_header: header.html
      after_body: reportdown.js
vignette: >
  %\VignetteIndexEntry{Multi-Omics Exploratory Analysis with cliomicplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7.5,
  fig.height = 5.5,
  fig.align  = "center"
)
```

```{r include-lightbox, echo=FALSE, results='asis'}
knitr::asis_output(paste(readLines('lightbox.html'), collapse='\n'))
```

## Overview

This vignette demonstrates a multi-omics exploratory analysis workflow using
**cliomicplot**. We'll generate ten figures — from core omics visualizations to
advanced analyses inspired by xOmicsShiny:

1. **Figure 1** — PCA with confidence ellipses
2. **Figure 2** — Correlation heatmap with clustering and significance
3. **Figure 3** — Volcano plot for differential expression
4. **Figure 4** — MA plot for DE quality assessment
5. **Figure 5** — Infographic bar chart
6. **Figure 6** — Loading real multi-omics data (xOmicsShiny format)
7. **Figure 7** — Expression pattern clustering (k-means/PAM/Mfuzz)
8. **Figure 8** — Rank-abundance curve with marginal distribution
9. **Figure 9** — DEG comparison scatter (two contrasts)
10. **Figure 10** — Diamond bubble heatmap (feature × condition matrix)

```{r}
library(cliomicplot)
clipar(stat.test = NULL)  # disable auto stat annotations
```

---

## Figure 1: PCA with Confidence Ellipses

Principal Component Analysis (PCA) is the workhorse of omics data exploration.
cliomicplot can compute PCA from raw data or accept pre-computed results:

### From raw data

```{r}
clitheme("nature")

cliplot(iris[, 1:4],
        type = type_pca(
          pc_x          = 1,
          pc_y          = 2,
          center        = TRUE,
          scale.        = TRUE,
          add_ellipse   = TRUE,
          ellipse_level = 0.95,
          point_size    = 2.5,
          label_samples = FALSE
        ),
        by       = iris$Species,
        palette  = "npg",
        stat.test = NULL,
        title    = "**Figure 1** — PCA of Iris Morphometric Data",
        subtitle = "PC1 (72.9%) vs. PC2 (22.9%), 95% CI ellipses") +
  cli_markdown()
```

The axis labels automatically include variance explained percentages, and
confidence ellipses are drawn around each group.

### From pre-computed PCA

When you already have PCA results (e.g., from `prcomp()`, `DESeq2`, or another
pipeline), cliomicplot accepts the scores directly:

```{r}
pca_res <- prcomp(iris[, 1:4], center = TRUE, scale. = TRUE)
scores  <- as.data.frame(pca_res$x)
scores$Species <- iris$Species

cliplot(PC2 ~ PC1, data = scores, by = scores$Species,
        type      = "points",
        stat.test = NULL,
        palette   = "d3_category10",
        title     = "**PCA Scores** — Pre-computed by prcomp()",
        theme     = "science") +
  cli_markdown()
```

### PCA with sample labels

```{r}
cliplot(iris[, 1:4],
        type = type_pca(
          add_ellipse   = FALSE,
          label_samples = TRUE,
          label_size    = 2.5,
          point_size    = 2
        ),
        by      = iris$Species,
        palette = "jco",
        stat.test = NULL,
        title   = "**PCA with Sample Labels**") +
  cli_markdown()
```

---

## Figure 2: Correlation Matrix

Correlation analysis with significance stars and hierarchical clustering:

```{r}
clitheme("cli_minimal")

cliplot(mtcars[, 1:7],
        type = type_correlation(
          method    = "pearson",
          type      = "lower",
          add_coef  = TRUE,
          coef_size = 3.5,
          cluster   = TRUE,
          sig_level = 0.001
        ),
        stat.test = NULL,
        title = "**Figure 2** — Pearson Correlation Matrix (mtcars)",
        subtitle = "Lower triangle | *** p < 0.001") +
  cli_markdown()
```

### Correlation options

```{r}
# Spearman correlation, upper triangle, no coefficients
cliplot(mtcars[, 1:5],
        type = type_correlation(
          method   = "spearman",
          type     = "upper",
          add_coef = FALSE,
          cluster  = FALSE
        ),
        title = "**Spearman Correlation** — Upper Triangle") +
  cli_markdown()
```

```{r}
# Full matrix with all significance levels
cliplot(mtcars[, 1:6],
        type = type_correlation(
          method    = "pearson",
          type      = "full",
          add_coef  = TRUE,
          cluster   = TRUE,
          sig_level = 0.05
        ),
        title = "**Full Correlation Matrix** with Significance Stars") +
  cli_markdown()
```

---

## Figure 3: Volcano Plot — Differential Expression

The volcano plot is the canonical visualization for differential expression analysis.
We'll generate realistic simulated data:

```{r}
set.seed(42)

# Generate 8000 "genes" — most are null
n_genes <- 8000
de_data <- data.frame(
  logFC   = rnorm(n_genes, 0, 0.7),
  PValue  = 10^-rnorm(n_genes, 2.5, 1.8),
  row.names = paste0("GENE", 1:n_genes)
)

# Inject 200 "true" DE genes
sig_idx <- sample(1:n_genes, 200)
de_data$logFC[sig_idx]  <- rnorm(200, ifelse(runif(200) > 0.5, 2.0, -2.0), 0.5)
de_data$PValue[sig_idx] <- 10^-runif(200, 5, 15)

cat(sprintf("Genes tested: %d\n", n_genes))
cat(sprintf("DE genes (|log2FC| > 1, p < 0.05): %d\n",
            sum(abs(de_data$logFC) > 1 & de_data$PValue < 0.05)))
```

```{r}
clitheme("cell")

cliplot(-log10(PValue) ~ logFC, data = de_data,
        type = type_volcano(
          pval_cutoff  = 0.01,
          fc_cutoff    = 1,
          label_genes  = "significant",
          max_overlaps = 25,
          point_alpha  = 0.5
        ),
        title    = "**Figure 3** — Volcano Plot of Differential Expression",
        subtitle = "FDR < 0.01, |log<sub>2</sub>FC| > 1",
        caption  = paste0("Top labeled genes; ", n_genes, " probes tested")) +
  cli_markdown()
```

### Volcano plot color scheme

| Category | Condition | Default Color |
|---|---|---|
| **Up** (red) | $p < p_\text{cutoff}$ and $\log_2(\text{FC}) > \text{fc\_cutoff}$ | `#E64B35` |
| **Down** (blue) | $p < p_\text{cutoff}$ and $\log_2(\text{FC}) < -\text{fc\_cutoff}$ | `#4DBBD5` |
| **NS** (grey) | All other genes | `grey70` |

### Customizing the volcano

```{r}
# Tighter thresholds, different colors
cliplot(-log10(PValue) ~ logFC, data = de_data,
        type = type_volcano(
          pval_cutoff  = 0.001,
          fc_cutoff    = 1.5,
          label_genes  = NULL,       # no labels
          up_color     = "#BC3C29",  # NEJM red
          down_color   = "#0072B5",  # NEJM blue
          ns_color     = "grey85",
          point_size   = 1.5,
          point_alpha  = 0.4
        ),
        palette = "volcano",
        title   = "**Volcano Plot** — Stringent Thresholds",
        subtitle = "p < 0.001, |log<sub>2</sub>FC| > 1.5") +
  cli_markdown()
```

---

## Figure 4: MA Plot — DE Quality Check

The MA (Mean-Average) plot complements the volcano by showing the relationship
between fold-change and mean expression level:

```{r}
set.seed(123)
ma_data <- data.frame(
  baseMean       = 10^runif(3000, 0, 5),
  log2FoldChange = rnorm(3000, 0, 0.6),
  padj           = runif(3000)
)
# Add "real" DE genes
ma_data$log2FoldChange[sample(1:3000, 150)] <- rnorm(150, 2, 0.5)
ma_data$log2FoldChange[sample(1:3000, 150)] <- rnorm(150, -2, 0.5)

clitheme("science")

cliplot(log2FoldChange ~ baseMean, data = ma_data,
        type = type_ma(
          pval_cutoff = 0.05,
          add_loess   = TRUE,
          loess_color = "#3C5488",
          sig_color   = "#E64B35",
          ns_color    = "grey70",
          point_size  = 1.2,
          point_alpha = 0.4
        ),
        title    = "**Figure 4** — MA Plot",
        subtitle = "DE genes highlighted (padj < 0.05), LOESS trend") +
  cli_markdown()
```

The MA plot shows:
- **y-axis**: $\log_2$ fold change (biological effect size)
- **x-axis**: Mean normalized counts (expression level, log₁₀ scale)
- **Red points**: Significant DE genes
- **Blue curve**: LOESS trend line (checks for intensity-dependent bias)
- **Horizontal dashed line**: $\log_2\text{FC} = 0$ (no change)

---

## Heatmap: Clustered Expression Matrix

For completeness, here's a clustered heatmap workflow:

```{r}
set.seed(789)
expr_mat <- matrix(rnorm(500), nrow = 50)
rownames(expr_mat) <- paste0("Gene_", 1:50)
colnames(expr_mat) <- paste0("Sample_", sprintf("%02d", 1:10))

# Column annotations
ann_col <- data.frame(
  Condition = rep(c("Control", "Treatment"), each = 5),
  Batch     = rep(c("A", "B"), 5),
  row.names = colnames(expr_mat)
)

# Annotation color mapping
ann_colors <- list(
  Condition = c(Control = "#4DBBD5", Treatment = "#E64B35"),
  Batch     = c(A = "#E18727", B = "#3C5488")
)
```

```{r, eval=requireNamespace("ComplexHeatmap", quietly=TRUE)}
cliplot(expr_mat,
        type = type_heatmap(
          scale             = "row",
          cluster_rows      = TRUE,
          cluster_cols      = TRUE,
          annotation_col    = ann_col,
          annotation_colors = ann_colors,
          color_low         = "#4575B4",
          color_mid         = "white",
          color_high        = "#D73027",
          fontsize          = 9
        ))
```

> **Note:** The heatmap uses `ComplexHeatmap` when available, falling back to
> `pheatmap`. Install with `install.packages("ComplexHeatmap")` for the best experience.

---

## Figure 5: Infographic Bar Chart — Results Comparison

The `type_infobar()` creates media-ready bar charts with per-category custom
colours, footer boxes, and reference lines. Use it for polling, survey results,
or any side-by-side comparison where each category carries semantic colouring:

```{r}
infobar_data <- data.frame(
  endpoint  = c("ORR", "DCR", "PFS6", "OS12", "AE3"),
  value     = c(42, 68, 55, 72, 12),
  bar_color = c("#0073C2", "#4DBBD5", "#00A087", "#3C5488", "#E64B35"),
  box_color = c("#005A9E", "#3AA8C2", "#00806A", "#2A3E6E", "#C9302C"),
  group     = c("Efficacy", "Efficacy", "Efficacy", "Efficacy", "Safety"),
  change    = c("(+3)", "(+5)", "(-1)", "(+2)", "(-4)"),
  stringsAsFactors = FALSE
)

cliplot(value ~ endpoint, data = infobar_data,
        type = type_infobar(
          bar_colors      = setNames(infobar_data$bar_color,
                                     infobar_data$endpoint),
          box_colors      = setNames(infobar_data$box_color,
                                     infobar_data$endpoint),
          box_labels      = infobar_data$endpoint,
          box_sub_labels  = infobar_data$group,
          change_labels   = infobar_data$change,
          reference_line  = 40,
          reference_label = "40%",
          value_suffix    = "%",
          bar_width       = 0.82
        ),
        title    = "**Figure 5** — Efficacy & Safety Endpoint Comparison",
        subtitle = "Clinical trial results by endpoint category")
```

### Customization

| Parameter | Default | Description |
|---|---|---|
| `bar_colors` | `NULL` (jco palette) | Named vector of per-bar fill colours |
| `box_colors` | `NULL` (no boxes) | Named vector of footer box fill colours |
| `box_labels` | `NULL` | Labels inside footer boxes (vector, one per bar) |
| `box_sub_labels` | `NULL` | Sub-labels below footer boxes |
| `change_labels` | `NULL` | Change annotations below value labels |
| `reference_line` | `NULL` | y-axis value for a horizontal reference line |
| `reference_label` | `NULL` | Label text next to the reference line |
| `value_suffix` | `""` | Suffix appended to value labels (e.g. `"%"`) |

---

## Figure 6: Loading Real Multi-Omics Data

cliomicplot can work directly with multi-omics datasets from platforms like
**xOmicsShiny**. Here we demonstrate loading a real metabolomics dataset:

```{r}
# Load xOmicsShiny-format RData file
data_path = system.file("extdata", "AgingHFCD_Metabolomics.RData",
                         package = "cliomicplot")
if (file.exists(data_path)) {
  load(data_path)
  cat("Available datasets:", paste(names(DataInSets), collapse = ", "), "\n")
}
```

### Understanding xOmicsShiny data structure

xOmicsShiny datasets are organized as nested lists within `DataInSets`:

```{r}
if (exists("DataInSets") && length(DataInSets) > 0) {
  ds = DataInSets[[1]]
  cat("Components:", paste(names(ds), collapse = ", "), "\n")
  cat("Groups:", paste(ds$groups, collapse = ", "), "\n")
  cat("N samples:", length(ds$samples), "\n")
  cat("N features in data_long:", nrow(ds$data_long), "\n")
  if (!is.null(ds$results_long)) {
    cat("N tests:", nrow(ds$results_long), "\n")
    print(head(ds$results_long[, 1:6], 3))
  }
}
```

| Component | Description |
|---|---|
| `data_long` | Tidy expression data (feature × sample × group) |
| `data_wide` | Wide matrix (features × samples) |
| `results_long` | Differential expression / statistics results |
| `MetaData` | Sample-level metadata (group assignments, covariates) |
| `ProteinGeneName` | Feature identifier mapping table |

---

## Figure 7: Expression Pattern Clustering

Inspired by **xOmicsShiny's pattern module**, `type_pattern()` groups features
(genes, proteins, metabolites) by their expression profiles across conditions or
time points. This reveals coordinated regulation patterns — essential for
time-series and multi-condition omics studies.

We'll generate realistic time-course data for demonstration:

```{r}
set.seed(42)
n_genes  = 600
n_points = 7

time_points = paste0("T", 0:6)

# Generate cluster-specific expression patterns
make_cluster = function(n, pattern, noise_sd = 0.5) {
  base = switch(pattern,
    "up"      = seq(0, 2.5, length.out = n_points),
    "down"    = seq(2.0, -0.5, length.out = n_points),
    "early_up"= c(0, 0.3, 1.5, 2.2, 1.8, 1.2, 0.8),
    "late_up" = c(0, -0.2, -0.1, 0.1, 0.8, 1.8, 2.5),
    "transient"=c(0, 0.2, 1.8, 0.5, -0.3, -0.2, -0.1),
    "osc"     = c(0, 1.2, -0.3, 0.8, -0.4, 0.6, 0)
  )
  t(replicate(n, base + rnorm(n_points, 0, noise_sd)))
}

clusters = rbind(
  make_cluster(100, "up"), make_cluster(100, "down"),
  make_cluster(100, "early_up"), make_cluster(100, "late_up"),
  make_cluster(100, "transient"), make_cluster(100, "osc")
)
rownames(clusters) = paste0("Gene_", 1:nrow(clusters))
colnames(clusters) = time_points

dim(clusters)
```

```{r}
cliplot(clusters,
        type = type_pattern(
          k         = 6,
          method    = "kmeans",
          ncol      = 3,
          line_alpha = 0.3
        ),
        stat.test = NULL,
        title    = "**Figure 7** — Expression Pattern Clustering",
        subtitle = "K-means clustering of time-course profiles (n = 600 genes)") +
  cli_markdown()
```

### Pattern clustering methods

| Method | Description | Best for |
|---|---|---|
| `kmeans` | Hard clustering by Euclidean distance | General purpose, fast |
| `pam` | Partitioning Around Medoids | Robust to outliers |
| `mfuzz` | Soft (fuzzy) clustering via Mfuzz | Graded membership patterns |

```{r}
# Alternative: PAM clustering
cliplot(clusters,
        type = type_pattern(
          k      = 4,
          method = "pam",
          ncol   = 2
        ),
        stat.test = NULL,
        title    = "**Pattern Clustering** — PAM Method (k = 4)") +
  cli_markdown()
```

---

## Figure 8: Rank Abundance Curve

The rank-abundance curve (inspired by **xOmicsShiny's S-curve**) sorts all
features by their abundance and plots them against rank. It's a standard QC
plot in proteomics and metabolomics, revealing:

- **Dynamic range** of the assay
- **Distribution shape** (normal, log-normal, bimodal)
- **Outlier features** with extreme abundance

```{r}
set.seed(123)
n_features = 5000

# Simulate realistic proteomics intensity distribution
ra_data = data.frame(
  Feature   = paste0("PROT_", 1:n_features),
  Intensity = 10^rnorm(n_features, mean = 6.5, sd = 1.2),
  stringsAsFactors = FALSE
)
# Sort by intensity descending and add rank
ra_data = ra_data[order(-ra_data$Intensity), ]
ra_data$Rank = 1:n_features

# Highlight some known proteins
highlight = paste0("PROT_", c(10, 50, 200, 800, 3000))

head(ra_data)
```

```{r}
clitheme("science")

cliplot(Intensity ~ Rank, data = ra_data,
        type = type_rankabundance(
          log_scale      = "10",
          point_color    = "#4DBBD5",
          marginal_type  = "density",
          marginal_color = "#3C5488",
          label_genes    = highlight,
          label_color    = "#E64B35"
        ),
        stat.test = NULL,
        title    = "**Figure 8** — Rank-Abundance Curve",
        subtitle = "Simulated proteomics — log₁₀ intensity, top 5 proteins labeled",
        caption  = paste0(n_features, " features; log₁₀ scale")) +
  cli_markdown()
```

> **Tip:** Add a marginal distribution with `ggExtra::ggMarginal(p, type = "density")`
> for publication-quality rank-abundance plots with distribution annotation.

### Customizing the rank-abundance curve

```{r}
# No labels, with density marginal distribution
cliplot(Intensity ~ Rank, data = ra_data,
        type = type_rankabundance(
          log_scale     = "2",
          point_color   = "#0073C2",
          marginal_type = "density"
        ),
        stat.test = NULL,
        title = "**Rank-Abundance** — log₂ Scale, No Labels") +
  cli_markdown()
```

---

## Figure 9: DEG Comparison Scatter

Inspired by **xOmicsShiny's "DEGs in Two Comparisons"**, `type_deg_compare()`
plots the log₂ fold-changes of two independent contrasts against each other.
This reveals:

- **Concordant regulation** (both up or both down — top-right/bottom-left quadrants)
- **Discordant regulation** (opposite direction — top-left/bottom-right)
- **Comparison-specific effects** (significant in one but not the other)

```{r}
set.seed(789)
n_de = 2000

# Simulate two DE comparisons with shared and unique signals
shared_genes = sample(1:n_de, 300)
fc1 = rnorm(n_de, 0, 0.8)
fc2 = rnorm(n_de, 0, 0.8)

# Inject shared DE signal (concordant)
fc1[shared_genes] = fc1[shared_genes] + rnorm(300, 2.0, 0.5)
fc2[shared_genes] = fc2[shared_genes] + rnorm(300, 2.0, 0.5)

# Inject comparison-specific signals
fc1[sample(1:n_de, 80)] = rnorm(80, 2.5, 0.4)
fc2[sample(1:n_de, 80)] = rnorm(80, -2.5, 0.4)

de1 = data.frame(
  Gene   = paste0("GENE_", 1:n_de),
  logFC  = fc1,
  PValue = 10^-runif(n_de, 1, 8),
  stringsAsFactors = FALSE
)
de2 = data.frame(
  Gene   = paste0("GENE_", 1:n_de),
  logFC  = fc2,
  PValue = 10^-runif(n_de, 1, 8),
  stringsAsFactors = FALSE
)

# Label top genes from both comparisons
top_genes = paste0("GENE_", shared_genes[1:10])
```

```{r}
clitheme("cell")

cliplot(logFC ~ Gene, data = de1,
        type = type_deg_compare(
          comp2_data  = de2,
          merge_by    = "Gene",
          add_lm      = TRUE,
          label_genes = top_genes,
          same_scale  = TRUE
        ),
        stat.test = NULL,
        title    = "**Figure 9** — DEG Comparison Scatter",
        subtitle = "log₂FC: Comparison 1 (WT vs KO) × Comparison 2 (Treated vs Control)") +
  cli_markdown()
```

### Interpreting the quadrants

| Quadrant | Interpretation |
|---|---|
| **Top-right** (+,+) | Consistently upregulated in both comparisons |
| **Bottom-left** (−,−) | Consistently downregulated in both comparisons |
| **Top-left** (−,+) | Opposing direction — comparison-specific upregulation |
| **Bottom-right** (+,−) | Opposing direction — comparison-specific downregulation |

The linear regression line indicates the overall concordance between the two
comparisons (slope ≈ 1 = strong agreement).

---

## Figure 10: Diamond Bubble Heatmap

The `type_bubble_heatmap()` creates a matrix of diamond-shaped bubbles where
feature–sample combinations are encoded by **size** (e.g., fold-change) and
**colour** (e.g., significance). It's ideal for pathway × comparison matrices
or multi-marker panels, inspired by **xOmicsShiny's heatmap visualizations**:

```{r}
set.seed(456)
markers = paste0("Marker_", LETTERS[1:12])
cell_types = paste0("CellType_", 1:8)

bh_data = expand.grid(
  Marker = markers,
  CellType = cell_types,
  stringsAsFactors = FALSE
)
bh_data$FoldChange = round(runif(nrow(bh_data), 0.2, 8.6), 2)
bh_data$Padj       = runif(nrow(bh_data), 1e-10, 0.5)

# Make some markers stand out
bh_data$FoldChange[bh_data$Marker %in% c("Marker_A", "Marker_D", "Marker_H")] =
  bh_data$FoldChange[bh_data$Marker %in% c("Marker_A", "Marker_D", "Marker_H")] * 1.8
```

```{r}
cliplot(Marker ~ CellType, data = bh_data,
        by = bh_data$FoldChange,
        type = type_bubble_heatmap(
          label_threshold = 3.5,
          max_size        = 14,
          size_name       = "Fold Change",
          fill_name       = expression(-log[10]~"adjusted P"),
          fill_low        = "#4575B4",
          fill_mid        = "#FFFFBF",
          fill_high       = "#D73027"
        ),
        extra_data = list(fill = -log10(bh_data$Padj)),
        stat.test  = NULL,
        title      = "**Figure 10** — Diamond Bubble Heatmap",
        subtitle   = "Marker × Cell Type — size = fold-change, colour = significance") +
  cli_markdown()
```

### Bubble heatmap customization

| Parameter | Default | Description |
|---|---|---|
| `label_threshold` | `2.0` | Minimum size for in-cell value labels |
| `max_size` | `15` | Maximum bubble point size |
| `band_alpha` | `0.6` | Opacity for alternating row highlight bands |
| `fill_low/mid/high` | 3-colour gradient | Fill gradient stops for bubble fill |
| `size_name` | `"Fold change"` | Legend title for size scale |

---

## Summary

In this vignette we produced ten publication-ready omics figures:

1. **PCA** — Dimensionality reduction with confidence ellipses
2. **Correlation Matrix** — Pairwise correlations with hierarchical clustering
3. **Volcano Plot** — Differential expression significance vs. fold-change
4. **MA Plot** — Fold-change vs. mean expression with LOESS trend
5. **Infobar** — Per-category infographic comparison
6. **Real Multi-Omics Data** — Loading and exploring xOmicsShiny datasets
7. **Pattern Clustering** — K-means/PAM expression profile clustering
8. **Rank-Abundance Curve** — Feature abundance with marginal distributions
9. **DEG Comparison** — Concordance of fold-changes across two contrasts
10. **Diamond Bubble Heatmap** — Feature × condition matrix with dual encoding

Reset global settings:

```{r}
clitheme()
```
