---
title: "End-to-End Oncology Analysis with cliomicplot"
author: "Lucas VHH Tran"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    mathjax: null
    css: reportdown.css
    self_contained: true
    includes:
      in_header: header.html
      after_body: reportdown.js
vignette: >
  %\VignetteIndexEntry{End-to-End Oncology Analysis with cliomicplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 8,
  fig.height = 5.5,
  fig.align  = "center"
)
```

```{r include-lightbox, echo=FALSE, results='asis'}
knitr::asis_output(paste(readLines('lightbox.html'), collapse='\n'))
```

## Overview

This vignette demonstrates a complete oncology clinical & translational data
analysis workflow using **cliomicplot**. We'll generate publication-quality
figures commonly found in oncology manuscripts, with additional analyses
inspired by **xOmicsShiny**:

1. **Figure 1** — Kaplan-Meier overall survival curves
2. **Figure 2** — Forest plot from multivariable Cox regression
3. **Figure 3A** — Waterfall plot of best tumor response
4. **Figure 3B** — Swimmer plot of patient treatment courses
5. **Figure 4A** — Dose-response curves for drug sensitivity
6. **Figure 4B** — Gene expression time course trajectories
7. **Figure 4C** — Prism-style boxplot with P-value annotations

```{r}
library(cliomicplot)
library(survival)

# Set global defaults for this analysis
clipar(
  palette.qualitative = "nejm",
  stat.test           = NULL,
  file.res            = 600
)
```

---

## Figure 1: Kaplan-Meier Survival Curves

We use the `lung` dataset from the **survival** package to visualize overall survival
stratified by sex.

```{r}
clitheme("nejm")

cliplot(Surv(time, status) ~ sex, data = lung,
        type = type_km(
          risk_table  = TRUE,
          pval        = TRUE,
          conf_int    = TRUE,
          median_line = TRUE
        ),
        palette  = "nejm",
        title    = "**Figure 1** — Overall Survival by Sex",
        subtitle = "NCCTG Lung Cancer Trial (n = 228)",
        xlab     = "Time (days)",
        ylab     = "Survival Probability",
        legend   = "top") +
  cli_markdown()
```

### Key features demonstrated:

- `type_km()` with risk table, p-value, confidence intervals, and median survival line
- `clitheme("nejm")` for NEJM-style typography
- `cli_markdown()` for bold formatting in the title

### Customizing KM plots

```{r}
# Without risk table, cleaner look
cliplot(Surv(time, status) ~ sex, data = lung,
        type = type_km(risk_table = FALSE, conf_int = FALSE),
        palette = "lancet",
        theme   = "lancet",
        title   = "**Overall Survival** — Sex-Stratified") +
  cli_markdown()
```

---

## Figure 2: Forest Plot — Multivariable Cox Regression

Fit a Cox proportional hazards model and visualize the hazard ratios:

```{r}
# Fit Cox model
cox_fit <- coxph(Surv(time, status) ~ age + sex + ph.ecog + wt.loss, data = lung)
cox_sum <- summary(cox_fit)

# Extract results into a forest-plot-ready data frame
forest_df <- data.frame(
  Variable = c("Age (per year)", "Male vs. Female",
               "ECOG PS (per point)", "Weight loss (per lb)"),
  HR       = cox_sum$conf.int[, "exp(coef)"],
  CI_low   = cox_sum$conf.int[, "lower .95"],
  CI_high  = cox_sum$conf.int[, "upper .95"],
  P        = cox_sum$coefficients[, "Pr(>|z|)"]
)

forest_df
```

```{r}
clitheme("lancet")

cliplot(HR ~ Variable, data = forest_df,
        type = type_forest(
          ref_line   = TRUE,
          sort       = TRUE,
          point_size = 4,
          ci_width   = 1.2,
          sig_color  = "#E64B35",
          ns_color   = "#3C5488"
        ),
        title    = "**Figure 2** — Multivariable Cox Regression",
        subtitle = "Overall Survival — Hazard Ratios (95% CI)",
        caption  = "Lung Cancer Dataset (n = 228)") +
  cli_markdown()
```

### Interpreting the forest plot

- **Red points** = p < 0.05 (statistically significant)
- **Blue points** = p >= 0.05 (not significant)
- **Vertical dashed line** at HR = 1 (null effect)
- HR > 1 = increased risk; HR < 1 = protective effect

---

## Figure 3A: Waterfall Plot — Tumor Response

Generate a simulated immunotherapy trial dataset and visualize best tumor response:

```{r}
set.seed(42)
n_pts <- 30

imtrial <- data.frame(
  Patient    = paste0("IM-", sprintf("%03d", 1:n_pts)),
  BestChange = sort(rnorm(n_pts, mean = -22, sd = 32)),
  BestResp   = sample(c("CR", "PR", "SD", "PD"), n_pts,
                      replace = TRUE,
                      prob = c(0.07, 0.30, 0.36, 0.27)),
  Duration   = pmax(1, round(rnorm(n_pts, mean = 14, sd = 7), 1)),
  Arm        = sample(c("Anti-PD1", "Chemotherapy"), n_pts, replace = TRUE)
)

# Show first few rows
head(imtrial)
```

```{r}
clitheme("broadsheet")

cliplot(BestChange ~ Patient, data = imtrial,
        type = type_waterfall(
          show_labels = TRUE,
          label_size  = 2.5,
          bar_width   = 0.75,
          response_thresholds = c("PD" = 20, "SD" = -30)
        ),
        title    = "**Figure 3A** — Best Tumor Response (Waterfall)",
        subtitle = "RECIST v1.1 — % Change from Baseline",
        ylab     = "Best % Change from Baseline",
        legend   = "bottom") +
  cli_markdown()
```

### Response categories

| Category | Threshold | Color |
|---|---|---|
| CR (Complete Response) | Disappearance of all lesions | 🟢 Green |
| PR (Partial Response) | ≥ 30% decrease | 🔵 Blue |
| SD (Stable Disease) | Between +20% and −30% | 🟠 Orange |
| PD (Progressive Disease) | ≥ 20% increase | 🔴 Red |

---

## Figure 3B: Swimmer Plot — Patient Timelines

Add clinical event markers to the patient timeline:

```{r}
set.seed(123)
n_events_per_pt <- sample(1:3, n_pts, replace = TRUE)
n_total_events  <- sum(n_events_per_pt)

events_df <- data.frame(
  ID    = rep(imtrial$Patient, times = n_events_per_pt),
  Time  = round(runif(n_total_events, 0, 28), 1),
  Event = sample(c("Progression", "AE Gr≥3", "Dose Reduction", "Response"),
                 n_total_events,
                 replace = TRUE, prob = c(0.3, 0.25, 0.2, 0.25))
)

head(events_df)
```

```{r}
cliplot(Duration ~ Patient, data = imtrial,
        type = type_swimmer(
          bar_fill    = "BestResp",
          event_times = events_df,
          bar_alpha   = 0.85
        ),
        palette = "jco",
        title   = "**Figure 3B** — Patient Treatment Course (Swimmer)",
        subtitle = "Duration on study with key clinical events",
        xlab    = "",
        ylab    = "Time on Study (months)",
        legend  = "bottom") +
  cli_markdown()
```

### Event shape legend

| Event | Shape | Color |
|---|---|---|
| Progression | ✳ (asterisk, 8) | 🔴 Red |
| AE Gr≥3 | ● (circle, 16) | 🟠 Orange |
| Dose Reduction | ▲ (triangle, 17) | 🟤 Brown |
| Response | ◆ (diamond, 18) | 🟢 Green |
| Death | ✖ (cross, 4) | ⚫ Black |

---

## Figure 4A: Dose-Response Curves — Drug Sensitivity

Inspired by **xOmicsShiny's dose-response/omics module**, dose-response analysis
is essential in oncology for characterizing drug sensitivity and resistance.
Using `type_lines()` with confidence bands, we can visualize concentration-response
relationships across compounds:

```{r}
set.seed(42)

# Simulate an in vitro drug screen across 4 compounds
doses     = c(0.001, 0.01, 0.1, 1, 10, 100)
compounds = c("Paclitaxel", "Doxorubicin", "Cisplatin", "Erlotinib")

dr_data = do.call(rbind, lapply(1:4, function(i) {
  ic50 = c(0.08, 0.5, 2.0, 25)[i]
  hill = c(-1.2, -0.9, -1.5, -1.1)[i]
  max_inhib = c(95, 88, 92, 78)[i]
  resp = max_inhib / (1 + (doses / ic50)^(-hill)) +
         rnorm(length(doses), 0, 3)
  data.frame(
    Compound = compounds[i],
    Dose     = doses,
    Response = pmax(0, pmin(100, resp)),
    stringsAsFactors = FALSE
  )
}))

head(dr_data, 8)
```

```{r}
clitheme("science")

cliplot(Response ~ Dose | Compound, data = dr_data,
        type = type_lines(
          add_points  = TRUE,
          point_size  = 2.2,
          add_ribbon  = FALSE,
          group       = "Compound"
        ),
        palette = "jco",
        title   = "**Figure 4A** — In Vitro Drug Sensitivity Screen",
        subtitle = "Concentration–response curves (n = 4 compounds, 6 doses)",
        xlab    = expression("Concentration ["*mu*"M]"),
        ylab    = "Cell Viability (%)") +
  cli_markdown()
```

> **Note:** Dose-response data from high-throughput screens (e.g., xOmicsShiny)
> can be directly plotted with `cliplot()`. Use log-transformed x-axis with
> `ggplot2::scale_x_log10()` for standard pharmacology conventions.

```{r}
# Log-scale x-axis (standard pharmacology convention)
cliplot(Response ~ Dose | Compound, data = dr_data,
        type = type_lines(
          add_points  = TRUE,
          add_ribbon  = FALSE,
          group       = "Compound"
        ),
        palette = "npg",
        title   = "**Dose-Response** — Log₁₀ Concentration Scale",
        theme   = "cli_minimal",
        xlab    = expression("log"[10]*" Concentration ["*mu*"M]"),
        ylab    = "Cell Viability (%)") +
  cli_markdown() +
  ggplot2::scale_x_log10()
```

---

## Figure 4B: Gene Expression Time Course

Multi-condition or time-course gene expression profiles are a core analysis in
xOmicsShiny. Using `type_lines()` with facets, we can visualize longitudinal
biomarker trajectories across treatment arms:

```{r}
set.seed(789)
n_biomarkers = 8
biomarkers   = c("Ki-67", "p53", "BCL-2", "EGFR", "VEGF", "PD-L1", "HER2", "c-MYC")
timepoints   = c("Baseline", "Cycle 2", "Cycle 4", "Cycle 6", "EOT")

# Generate realistic fold-change trajectories
tc_data = do.call(rbind, lapply(1:8, function(i) {
  # Each biomarker has its own trajectory shape
  shapes = list(
    c(0, -0.3, -0.8, -1.5, -2.0),  # Ki-67: steady down
    c(0, 0.5, 0.8, 0.3, -0.2),     # p53: early up then down
    c(0, 0.2, -0.5, -1.2, -1.8),   # BCL-2
    c(0, -0.8, -1.5, -2.0, -2.5),  # EGFR
    c(0, -0.2, -0.5, -0.8, -1.2),  # VEGF
    c(0, 1.0, 2.0, 1.5, 0.5),     # PD-L1: up then slight decline
    c(0, -0.5, -1.0, -1.5, -2.0), # HER2
    c(0, 0.3, 0.8, 1.2, 1.5)      # c-MYC: steady up
  )
  shape = shapes[[i]]
  data.frame(
    Biomarker = biomarkers[i],
    Timepoint = factor(timepoints, levels = timepoints),
    FoldChange = shape + rnorm(5, 0, 0.15),
    stringsAsFactors = FALSE
  )
}))

head(tc_data)
```

```{r}
clitheme("broadsheet")

cliplot(FoldChange ~ Timepoint | Biomarker, data = tc_data,
        type = type_lines(
          add_points  = TRUE,
          point_size  = 3,
          group       = "Biomarker",
          linewidth   = 1.1
        ),
        palette = "lancet",
        title   = "**Figure 4B** — Longitudinal Biomarker Trajectories",
        subtitle = "8 biomarkers × 5 time points, Tumor vs Baseline log₂FC",
        ylab    = expression("log"[2]~"Fold Change from Baseline"),
        legend  = "right") +
  cli_markdown()
```

---

## Figure 4C: Prism-Style Boxplot with P-Value Bars

Inspired by **xOmicsShiny's expression module**, which supports GraphPad
Prism styling via `ggprism`. The built-in `stat.test` parameter in
cliomicplot provides automatic significance annotations:

```{r}
clitheme("cli_minimal")

# Simulate a pharmacodynamic biomarker study
set.seed(456)
pd_data = data.frame(
  Biomarker = rep(c("Ki-67", "p21", "γH2AX", "c-PARP"), each = 24),
  Change    = c(
    rnorm(24, -35, 12),  # Ki-67: high inhibition
    rnorm(24,  28, 15),  # p21: induction
    rnorm(24,  45, 18),  # γH2AX: DNA damage marker
    rnorm(24,  52, 20)   # c-PARP: apoptosis marker
  ),
  Arm       = rep(c("Control", "Low Dose", "High Dose"), 32),
  stringsAsFactors = FALSE
)
pd_data$Arm = factor(pd_data$Arm, levels = c("Control", "Low Dose", "High Dose"))

head(pd_data)
```

```{r}
cliplot(Change ~ Arm | Biomarker, data = pd_data,
        type = type_boxplot(
          add_jitter = TRUE,
          jitter_width = 0.22,
          stat.test  = "anova",
          stat.label = "p.signif"
        ),
        palette = "jco",
        title   = "**Figure 4C** — Pharmacodynamic Biomarker Response",
        subtitle = "Boxplot with one-way ANOVA significance annotations",
        ylab    = "% Change from Baseline (D7)") +
  cli_markdown()
```

> **Tip:** For prism-style theming, install `ggprism` and apply
> `ggplot2::theme_prism()` on top of your cliomicplot.

---

## Combined Multi-Panel Figure

Use **patchwork** to combine the waterfall and swimmer plots:

```{r}
library(patchwork)

# Regenerate with consistent theme
p1 <- cliplot(BestChange ~ Patient, data = imtrial,
              type = type_waterfall(show_labels = FALSE, bar_width = 0.7),
              theme = "cli_minimal",
              title = "**A** — Waterfall") + cli_markdown()

p2 <- cliplot(Duration ~ Patient, data = imtrial,
              type = type_swimmer(bar_fill = "BestResp", event_times = events_df),
              theme = "cli_minimal",
              title = "**B** — Swimmer") + cli_markdown()

p1 + p2 +
  plot_annotation(
    title   = "**Figure 3** — Tumor Response Assessment",
    caption = "Simulated immunotherapy trial data",
    theme   = ggplot2::theme(plot.title = ggtext::element_markdown())
  )
```

---

## Saving Figures for Publication

```{r eval=FALSE}
# Save individual figures as high-resolution files
cliplot(Surv(time, status) ~ sex, data = lung,
        type = type_km(risk_table = TRUE),
        theme = "nejm", palette = "nejm",
        file = "fig1_km.pdf", width = 8, height = 6)

cliplot(HR ~ Variable, data = forest_df,
        type = type_forest(),
        theme = "lancet",
        file = "fig2_forest.pdf", width = 7, height = 4.5)

cliplot(BestChange ~ Patient, data = imtrial,
        type = type_waterfall(),
        theme = "broadsheet",
        file = "fig3_waterfall.pdf", width = 10, height = 5.5)
```

---

## Figure 4: Clinical Trials Pipeline — Stacked Bar Chart

The new `type_trials()` creates a horizontal stacked bar chart ideal for showing
trial pipelines, project portfolios, or any categorical breakdown within groups:

```{r}
# Simulated trial pipeline data
pipeline <- data.frame(
  company = rep(c("AstraZeneca", "Roche", "MSD", "Lilly", "Hengrui",
                  "Sanofi", "Novartis", "BMS", "J&J", "Pfizer", "AbbVie"),
                each = 5),
  phase   = rep(c("I", "I/II", "II", "II/III", "III"), 11),
  n       = c(4,3,24,0,32, 10,8,21,0,19, 4,2,30,1,15, 6,3,15,0,17,
              15,0,15,0,8, 7,0,14,0,16, 8,3,7,0,17, 7,5,14,1,6,
              6,2,7,0,14, 9,2,10,0,6, 4,3,4,0,15)
)

phase_colors <- c(
  "I"      = "#F8D95C", "I/II"   = "#D69B3C", "II"     = "#A7BA91",
  "II/III" = "#B76348", "III"    = "#123F2A"
)
phase_text <- c(
  "I"      = "#173E2B", "I/II"   = "#173E2B", "II"     = "#173E2B",
  "II/III" = "white",   "III"    = "#F6F1DE"
)

cliplot(n ~ company | phase, data = pipeline,
        type = type_trials(
          segment_colors      = phase_colors,
          segment_text_colors = phase_text,
          label_threshold     = 3
        ),
        title    = "**Figure 4** — Industry-Sponsored Clinical Trials by Phase",
        subtitle = "Companies with ≥ 25 completed trials, 2025",
        caption  = "Source: Citeline Annual Completed Clinical Trials Report")
```

### Customization

| Parameter | Default | Description |
|---|---|---|
| `segment_colors` | `NULL` (jco palette) | Named vector mapping segments to fill colours |
| `segment_text_colors` | `NULL` (auto dark) | Named vector mapping segments to label text colours |
| `bar_height` | `0.72` | Bar height as fraction of row spacing |
| `label_threshold` | `3` | Minimum segment width to show internal label |
| `show_totals` | `TRUE` | Show total count label to the right |

---

## Summary

In this vignette we produced four publication-ready oncology figures using
cliomicplot's specialized plot types. The key takeaways:

1. **`type_km()`** for survival curves with risk tables
2. **`type_forest()`** for Cox regression hazard ratios
3. **`type_waterfall()`** for tumor response visualization
4. **`type_swimmer()`** for patient-level timelines
5. **`type_trials()`** for clinical trial pipeline stacked bar charts
6. **Dose-response curves** with `type_lines()` for drug sensitivity screens
7. **Time-course trajectories** with automatic biomarker profiling
8. **Prism-style statistics** built into `type_boxplot()` via `stat.test`
9. **`clitheme()`** for journal-specific styling
10. **`cli_markdown()`** for rich text formatting

Reset global settings when done:

```{r}
clitheme()
clipar(palette.qualitative = "jco", stat.test = NULL)
```
