---
title: "Themes and Palettes in cliomicplot"
author: "Lucas VHH Tran"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    mathjax: null
vignette: >
  %\VignetteIndexEntry{Themes and Palettes in cliomicplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center"
)
```

## Overview

cliomicplot ships with 10 publication-quality themes and 55+ color palettes.
This vignette covers both systems in depth — how to select, customize, register,
and combine them for maximum impact.

```{r}
library(cliomicplot)
```

---

## Part 1: Themes

### Theme Catalog

| Theme | Base Size | Grid | Panel Border | Best For |
|---|---|---|---|---|
| `cli_bw` | 11 | Yes | Yes | General purpose (default) |
| `cli_classic` | 11 | Yes | No | Traditional R look |
| `cli_minimal` | 11 | Faint | No | Modern, clean |
| `nature` | 10 | No | No | Nature submissions |
| `science` | 10 | No | No | Science / AAAS |
| `nejm` | 10 | Subtle | No | NEJM style |
| `lancet` | 10 | Subtle | No | The Lancet |
| `cell` | 10 | No | No | Cell Press |
| `broadsheet` | 14 | Horizontal | Yes | Print / Newspaper |
| `dark` | 11 | Subtle | No | Presentations / Slides |

### Side-by-side theme comparison

```{r}
# Generate the same plot with different themes
data(iris)
p <- function(thm) {
  cliplot(Sepal.Length ~ Petal.Length | Species, data = iris,
          palette = "jco", theme = thm,
          stat.test = NULL,
          title = thm)
}
```

```{r fig.width=10, fig.height=8}
library(patchwork)

(p("cli_bw") + p("cli_minimal")) /
(p("nature") + p("nejm")) /
(p("lancet") + p("cell")) +
  plot_annotation(title = "Theme Comparison Across Journals")
```

### Persistent vs. Ephemeral Themes

```{r}
# ── Persistent: applies to ALL subsequent plots ────────────────────
clitheme("nature")

cliplot(mpg ~ wt, data = mtcars,
        title = "Plot 1 — Nature theme persists")

cliplot(len ~ dose, data = ToothGrowth, type = "boxplot",
        title = "Plot 2 — Still Nature theme")

clitheme()  # ← reset to default
```

```{r}
# ── Ephemeral: applies to ONE plot only ────────────────────────────
cliplot(mpg ~ wt, data = mtcars, theme = "nature",
        title = "This plot: Nature")

cliplot(Sepal.Length ~ Species, data = iris, type = "boxplot",
        theme = "dark",
        title = "This plot: Dark")

cliplot(mpg ~ wt, data = mtcars,
        title = "Back to default — cli_bw")
```

### Custom Theme Registration

Register your own themes for reuse across projects:

```{r}
# Register a custom theme
clitheme_register("my_submission",
  base = "cli_bw",
  axis.text     = ggplot2::element_text(size = 12, face = "bold"),
  plot.title    = ggplot2::element_text(size = 16, hjust = 0.5, face = "bold"),
  plot.subtitle = ggplot2::element_text(size = 11, hjust = 0.5, color = "grey40"),
  legend.position   = "bottom",
  legend.background = ggplot2::element_rect(color = "grey80")
)

# Use it
clitheme("my_submission")
cliplot(mpg ~ wt, data = mtcars,
        title = "Custom 'my_submission' Theme",
        subtitle = "12pt bold text, centered")
clitheme()
```

```{r}
# List all themes (including custom ones)
clitheme_list()

# Remove a custom theme
clitheme_unregister("my_submission")
```

### Theme Overrides

Pass inline overrides to tweak any theme element:

```{r}
# Start with nature, but override axis text size
clitheme("nature", axis.text = ggplot2::element_text(size = 14))
cliplot(mpg ~ wt, data = mtcars,
        title = "Nature theme with larger axis text")
clitheme()
```

---

## Part 2: Color Palettes

cliomicplot's 55+ palettes are organized into logical categories:

### Quick Lookup

```{r}
# List all palettes
all_pals <- cli_palette_list()
cat(sprintf("Total palettes: %d\n", length(all_pals)))
cat("First 25:\n")
cat(paste(head(all_pals, 25), collapse = ", "))
```

### Preview a Palette

```{r}
cli_palette_show("jco")
```

```{r fig.width=10, fig.height=12, eval=FALSE}
# Show all journal palettes side by side
journal_pals <- c("npg", "nejm", "lancet", "jama", "jco", "bmj", "frontiers")

par(mfrow = c(4, 2), mar = c(1, 3, 2, 1))
for (pal_name in journal_pals) {
  cols <- cliomicplot:::get_cli_palette(pal_name, 10)
  barplot(rep(1, length(cols)), col = cols, border = NA,
          main = pal_name, axes = FALSE, space = 0)
}
par(mfrow = c(1, 1))
```

### Palette Categories

#### 📰 Journal Palettes

```{r}
cliplot(Sepal.Length ~ Petal.Length | Species, data = iris,
        palette = "npg",
        title = "NPG (Nature Reviews Cancer)")

cliplot(Sepal.Length ~ Petal.Length | Species, data = iris,
        palette = "lancet",
        title = "The Lancet")
```

#### 🧬 Genomics & Bioinformatics

```{r}
# UCSC Genome Browser chromosome colors
cliplot(Sepal.Length ~ Petal.Length | Species, data = iris,
        palette = "ucscgb",
        title = "UCSC Genome Browser")

# COSMIC Cancer Hallmarks
cliplot(Sepal.Length ~ Petal.Length | Species, data = iris,
        palette = "cosmic",
        title = "COSMIC Hallmarks")
```

#### ♿ Accessibility (Colorblind-Safe)

These palettes are designed to be distinguishable under common forms of
color vision deficiency (deuteranopia, protanopia, tritanopia):

```{r}
cliplot(Sepal.Length ~ Petal.Length | Species, data = iris,
        palette = "okabe_ito",
        title = "Okabe & Ito (2008) — Colorblind-safe")

cliplot(Sepal.Length ~ Petal.Length | Species, data = iris,
        palette = "tol_muted",
        title = "Paul Tol — Muted, Colorblind-safe")
```

#### 🔥 Heatmap & Diverging

Use `type = "continuous"` with `palette_scale()` for gradient data:

```{r fig.width=8, fig.height=4}
library(ggplot2)

p1 <- ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_tile() +
  palette_scale("heatmap_rdbu", "fill", type = "continuous") +
  labs(title = "heatmap_rdbu") +
  theme_minimal() +
  theme(legend.position = "none")

p2 <- ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_tile() +
  palette_scale("heatmap_prgn", "fill", type = "continuous") +
  labs(title = "heatmap_prgn") +
  theme_minimal() +
  theme(legend.position = "none")

p1 + p2
```

#### 🌃 Dark / Presentation Palettes

```{r}
clitheme("dark")

cliplot(mpg ~ wt | factor(cyl), data = mtcars,
        palette = "neon",
        title = "Neon Palette on Dark Theme")

cliplot(Sepal.Length ~ Petal.Length | Species, data = iris,
        palette = "cyberpunk",
        title = "Cyberpunk Palette on Dark Theme")

clitheme()
```

### Using Palettes as Standalone Scales

All palettes work with **any ggplot2 plot** — not just cliomicplot:

```{r}
# Discrete colour scale
ggplot(iris, aes(Sepal.Length, Petal.Length, colour = Species)) +
  geom_point(size = 3) +
  palette_scale("jama", "color") +
  theme_bw() +
  labs(title = "JAMA palette via palette_scale()")

# Discrete fill scale
ggplot(mpg, aes(class, fill = class)) +
  geom_bar() +
  palette_scale("uchicago", "fill") +
  theme_classic() +
  labs(title = "UChicago palette via palette_scale()",
       subtitle = "Works on any ggplot2 plot!")
```

### Convenience Scale Wrappers

```{r}
# scale_color_cli() — discrete colour
ggplot(iris, aes(Sepal.Length, Petal.Length, color = Species)) +
  geom_point(size = 3) +
  scale_color_cli("jco") +
  theme_bw()

# scale_fill_cli_c() — continuous fill
ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_tile() +
  scale_fill_cli_c("blues") +
  theme_minimal()
```

### Reversing and Alpha

```{r}
# Reverse a palette
ggplot(iris, aes(Sepal.Length, Petal.Length, color = Species)) +
  geom_point(size = 3) +
  palette_scale("jco", "color", reverse = TRUE) +
  theme_bw() +
  labs(title = "JCO palette — reversed")

# Apply transparency
ggplot(iris, aes(Sepal.Length, Petal.Length, color = Species)) +
  geom_point(size = 3) +
  palette_scale("cosmic", "color", alpha = 0.5) +
  theme_bw() +
  labs(title = "COSMIC palette — 50% alpha")
```

---

## Part 3: Combining Themes and Palettes

The most powerful aspect of cliomicplot is how themes and palettes work together.
Here are some tested combinations for common use cases:

| Scenario | Theme | Palette |
|---|---|---|
| Nature submission | `nature` | `npg` |
| NEJM oncology paper | `nejm` | `nejm` |
| Lancet clinical trial | `lancet` | `lancet` |
| Cell genomics paper | `cell` | `cosmic` or `ucscgb` |
| Conference slides | `dark` | `neon` or `cyberpunk` |
| Preprint / bioRxiv | `broadsheet` | `jco` |
| General-purpose (safe) | `cli_bw` | `okabe_ito` |

```{r}
# Example: NEJM oncology paper
clitheme("nejm")
cliplot(Sepal.Length ~ Petal.Length | Species, data = iris,
        palette = "nejm",
        title = "NEJM Theme + NEJM Palette")
clitheme()
```

```{r}
# Example: Genomics paper with COSMIC
clitheme("cell")
cliplot(Sepal.Length ~ Petal.Length | Species, data = iris,
        palette = "cosmic",
        title = "Cell Theme + COSMIC Palette")
clitheme()
```

---

## Summary

- **10 themes** — from journal-specific to presentation-ready
- **55+ palettes** — journal, genomics, accessibility, design, and pop culture
- Themes can be **persistent** (`clitheme("nature")`) or **ephemeral** (`theme = "nature"`)
- Register custom themes with `clitheme_register()`
- All palettes work as standalone `palette_scale()` for any ggplot2 plot
- Combine themes and palettes freely for the perfect publication look

```{r}
# Reset to defaults
clitheme()
```
