---
title: "Volcano Plots — from Quick to Publication-Ready"
author: "Lucas VHH Tran"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    includes:
      in_header: header.html
vignette: >
  %\VignetteIndexEntry{Volcano Plots — from Quick to Publication-Ready}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 8,
  fig.height = 6,
  fig.align  = "center",
  message   = FALSE,
  warning   = FALSE
)
```

```{r include-lightbox, echo = FALSE, results = 'asis'}
knitr::asis_output(paste(readLines('lightbox.html'), collapse = '\n'))
```

```{r}
library(cliomicplot)
```

## Overview

The volcano plot is the canonical visualisation for differential-expression
analysis.  This vignette shows you the **one-liner** with `cliplot()` and then
dives into customisation — adjusted p-values, flexible fold-change thresholds,
gene labelling, and rich annotation call-outs — all with the
`type_volcano()` constructor.

---

## 1. Simulated data

Replace this section with your own DESeq2 / edgeR / limma output table.

```{r}
set.seed(2026)

de_results <- data.frame(
  gene    = paste0("Gene_", 1:1000),
  log2FC  = rnorm(1000, mean = 0, sd = 0.7),
  pvalue  = runif(1000, 0.00001, 1),
  padj    = p.adjust(runif(1000, 0.00001, 1), method = "BH"),
  row.names = paste0("Gene_", 1:1000)
)

# Inject a strongly significant hit
de_results <- rbind(
  de_results,
  data.frame(
    gene    = "CEACAM5",
    log2FC  = 2.79,
    pvalue  = 0.0001,
    padj    = 0.0027,
    row.names = "CEACAM5"
  )
)

head(de_results[, c("gene","log2FC","pvalue","padj")])
```

> **Your data.**  Columns must include a log~2~ fold-change, a (preferably
> adjusted) p-value, and optional gene identifiers as `row.names`.

---

## 2. Quick volcano — one call

```{r}
cliplot(-log10(padj) ~ log2FC, data = de_results,
        type  = type_volcano(
          pval_cutoff  = 0.05,
          fc_cutoff    = 0.25,
          point_alpha  = 0.65,
          label_genes  = "significant",
          max_overlaps = 20
        ),
        palette = "nature",
        theme   = "nature",
        title   = "Differential expression — quick volcano",
        subtitle = expression(FDR < 0.05 * ", |" * log[2] * "FC| > 0.25"))
```

That single call produces:

- Points coloured by significance class (**Up** / **Down** / **NS**)
- Dashed threshold lines at the chosen p-value and fold-change cutoffs
- Automatic `ggrepel` labels for every gene passing both thresholds
- A journal-style theme with curated palette

---

## 3. Customising every parameter

Every visual element is exposed.  Below we tighten the thresholds, change the
colour mapping, add a shaded significance region, and label only the top 15
hits by adjusted p-value.

```{r}
# Identify top-15 significant genes
sig_idx <- order(de_results$padj)[1:15]
top_genes <- rownames(de_results)[sig_idx]

cliplot(-log10(padj) ~ log2FC, data = de_results,
        type = type_volcano(
          pval_cutoff    = 0.01,
          fc_cutoff      = 0.50,
          up_color       = "#C95C43",
          down_color     = "#4C8BA8",
          ns_color       = "#B8B8B8",
          point_alpha    = 0.62,
          label_genes    = top_genes,
          max_overlaps   = 30,
          cutoff_region_alpha = 0.04,
          show_threshold_labels = TRUE
        ),
        palette = "jco",
        title   = "Stringent thresholds — top 15 genes labelled",
        subtitle = expression(FDR < 0.01 * ", |" * log[2] * "FC| > 0.50"))
```

---

## 4. Gene annotation call-out

For a publication figure you often want one or two genes highlighted with a
**detailed annotation box** reporting the fold-change, confidence interval, and
exact p-values.  `type_volcano` works with plain `geom_label_repel`; you can
layer a richer annotation on top using `ggtext::geom_richtext`.

```{r}
# Build annotation for CEACAM5
highlight <- de_results[rownames(de_results) == "CEACAM5", ]
highlight$ci_low   <- 1.86
highlight$ci_high  <- 3.73
highlight$annot <- paste0(
  "<b><i>", highlight$gene, "</i></b>",
  "<br><br>",
  "<b>Fold change = ", round(highlight$log2FC, 2), "</b>",
  " (95% CI: ", highlight$ci_low, ", ", highlight$ci_high, ")",
  "<br>",
  "<b>Unadjusted P-value</b> &lt; 0.0001",
  "<br>",
  "<b>Adjusted P-value = ", format(highlight$padj, digits = 2), "</b>"
)

# Base volcano from cliplot, then layer annotation
p_base <- cliplot(-log10(padj) ~ log2FC, data = de_results,
  type = type_volcano(
    pval_cutoff  = 0.05,
    fc_cutoff    = 0.25,
    label_genes  = "significant",
    max_overlaps = 15,
    point_alpha  = 0.60
  ),
  palette = "nature", theme = "nature",
  title   = "Volcano plot with gene annotation call-out",
  subtitle = expression(FDR < 0.05 * ", |" * log[2] * "FC| > 0.25"))

# Layer highlight + annotation on top of the returned ggplot
p_annot <- p_base +
  ggplot2::geom_point(
    data = highlight,
    ggplot2::aes(x = log2FC, y = -log10(padj)),
    inherit.aes = FALSE,
    shape = 21, size = 5, stroke = 1.4,
    fill  = "#C95C43", colour = "black"
  ) +
  ggtext::geom_richtext(
    data = highlight,
    ggplot2::aes(x = log2FC - 1.8, y = -log10(padj) + 1.8,
                 label = annot),
    inherit.aes = FALSE,
    hjust = 0, vjust = 0.5,
    size = 3.5, fill = "#F9E5D5", colour = "black",
    label.colour = NA, label.padding = grid::unit(c(8, 10, 8, 10), "pt")
  ) +
  ggplot2::geom_segment(
    data = highlight,
    ggplot2::aes(x = log2FC, y = -log10(padj),
                 xend = log2FC - 0.3, yend = -log10(padj) + 0.55),
    inherit.aes = FALSE,
    linewidth = 0.5, colour = "#555555"
  )

p_annot
```

---

## 5. Using your own data

Your differential-expression pipeline (DESeq2, edgeR, limma, …) produces a
table with at least these columns:

| Column | Description |
|---|---|
| `log2FC` | log~2~ fold-change |
| `pvalue` | raw p-value |
| `padj` | multiple-testing adjusted p-value (FDR) |

Row names should hold gene / probe identifiers.

```{r eval = FALSE}
# Example: loading DESeq2 output
res <- read.csv("differential_expression_results.csv")
rownames(res) <- res$gene_symbol

cliplot(-log10(padj) ~ log2FC, data = res,
        type = type_volcano(
          pval_cutoff  = 0.05,
          fc_cutoff    = 0.25,
          label_genes  = "significant",
          max_overlaps = 20
        ),
        title = "Differential expression results")
```

> **Tip.**  If your adjusted p-value column contains exact zeros (common with
> DESeq2 for highly significant genes), cap them before calling `cliplot`:
>
> ```{r eval = FALSE}
> res$padj <- pmax(res$padj, 1e-300)
> ```

---

## 6. Export for publication

```{r eval = FALSE}
# High-resolution PNG for manuscript
cliplot(-log10(padj) ~ log2FC, data = de_results,
        type = type_volcano(),
        theme = "nature", palette = "nature",
        file = "fig_volcano.png", width = 9, height = 7)

# Vector PDF for journal submission
cliplot(-log10(padj) ~ log2FC, data = de_results,
        type = type_volcano(),
        theme = "nature", palette = "nature",
        file = "fig_volcano.pdf", width = 9, height = 7)
```

---

## Summary

| What you want | How to do it |
|---|---|
| **Quick volcano** | `cliplot(-log10(padj) ~ log2FC, data, type = "volcano")` |
| **Custom thresholds** | `type_volcano(pval_cutoff = 0.01, fc_cutoff = 0.5)` |
| **Label all significant genes** | `type_volcano(label_genes = "significant")` |
| **Label specific genes** | `type_volcano(label_genes = c("TP53","EGFR","BRCA1"))` |
| **Change colours** | `type_volcano(up_color = "#FF0000", down_color = "#0000FF")` |
| **Annotation call-out** | Returned `ggplot` object + `ggtext::geom_richtext()` |
| **Export** | `file = "fig.pdf"`, `width = 9`, `height = 7` |
